SeedLing: Building and using a seed corpus \\ for the Human Language Project

### 2014-03-22 docs/computEL/data.tex
edits to intro, related work, and data

@@

According to Ethnologue, there are 7105 living languages, and 147 living language families. Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages. To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) in figure \ref{fig:heatmap}. The values in each cell have been colored according to proportion of languages represented, with green indicated good coverage and red poor. It's interesting to note that vigorous languages (6a) are poorly represented across all data sources, and worse than more endangered categories. In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.

@@

\begin{table}[h!]
\small
\centering
    \begin{tabular}{l|rr|rr}
    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
    \textbf{Combined}	& 1,347			 & 106 
    \end{tabular}
\caption{Corpus Coverage}
\label{table:corpus}
\end{table}

@@

Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}. Aligned texts consist of multiple parallel documents, aligned at the document, sentence, or word level. The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.  A monolingual document is simply one for which the document ID only has a single language code associated with it.  Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.

Analyses texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.  This is stored as a table, where each row represents a word, and each column a type of information.  They propose the following columns: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.

Out of our data sources, three can be straightforwardly represented in their aligned text structure.  However, ODIN contains richer annotations, which are in fact difficult to fit into their proposal, which we discuss in section \ref{sec:odin} below.

Although data size matters in general NLP, \emph{universality} is the utmost priority for a universal corpus. We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): The Online Database of Interlinear Text (ODIN), The Omniglot website, The Universal Declaration of Human Rights (UHDR), and Wikipedia.

Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN).  We describe each source in the following subsections.

\subsubsection{Universal Declaration of Human Rights}

\subsubsection{Wikipedia}

Wikipedia\footnote{url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability. At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages. Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.

\subsubsection{Omniglot}

The Omniglot website\footnote{url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages. We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages. The urls for these resources are:

\begin{itemize}[noitemsep]
\item \url{www.omniglot.com/language/phrases/*} 
\item \url{www.omniglot.com/babel/*}
\end{itemize}

\subsubsection{ODIN} \label{sec:odin}

ODIN (The Online Database of Interlinear Text) is a repository of interlinear glossed texts (IGTs) extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.  Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation.  An IGT consists of three lines: (i) the source, a sentence in a target language, (ii) the gloss, an analysis of each element in the source, and (iii) the translation, done at the sentence level. The gloss line can additionally include a number of linguistic terms, which means that the gloss can more properly described as being written in a metalanguage, rather than a natural language.  In ODIN, translations are into English, and glosses are written in an English-based metalanguage.  An accepted set of guidelines are given by the Leipzig Glossing Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}, where morphemes within words are separated by hyphens (or equal signs, for clitics), and the same number of hyphens should appear in each word of the source and gloss.

In contrast, the proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words. When we try to adapt IGTs into this format, we run into certain problems.  Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.  Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word). Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which undermines the central position of words in the proposed data structure.

Secondly, it is unclear how to represent the gloss.  Since the gloss line is written in a natural language, we cannot treat it as a simple translation.  However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.  We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".  Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.

One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.  However, this loses information, since we no longer no which morpheme has which meaning.  It would be ridiculous to throw away information in the IGTs, but to keep all information, we need to modify the \newcite{abney2011data}'s proposal.

Even though glosses in ODIN are based on English, having been extracted from English-language documents, this is not true of IGTs in general.  In particular, it is common for documentary linguists working on indigenous languages of the Americas to provide glosses and translations based on Spanish.  For this reason, we believe it would be wise to specific the language used to produce the gloss.  Since it is not quite the language itself, but a metalanguage, one solution would be to use new language codes that make it clear both that a metalanguage is being used, and also what natural language it is based on.  The five-letter code \texttt{gloss} cannot be confused with any code in any version of ISO 639 (which includes codes of length two to four).  Following the convention that subvarieties of a language are indicated with suffixes, we can append the code of the natural language being used.  For example, glosses into English and Spanish-based metalanguages would be given the codes \texttt{gloss-eng} and \texttt{gloss-spa}, respectively.

>>>>

Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}. Aligned texts consist of multiple parallel documents, aligned at the document, sentence, or word level. The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.  A monolingual document is simply one with only a single associated language code.  Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.

Analysed texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.  This is stored as a table, where each row represents a word, and each column a type of information.  They propose the following columns: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.

Out of our data sources, three can be straightforwardly represented in their aligned text structure.  However, ODIN contains richer annotations, which are in fact difficult to fit into their proposal, and which we discuss in section \ref{sec:odin} below.

Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus. We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.

Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN).  Table~\ref{table:corpus} gives some coverage statistics, and we describe each source in the following subsections.

\begin{table}[t]
\small
\centering
    \begin{tabular}{l|rr|rr}
    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
    \textbf{Combined}	& 1,347			 & 106 
    \end{tabular}
\caption{Corpus Coverage}
\label{table:corpus}
\end{table}

\paragraph{Universal Declaration of Human Rights.}

\paragraph{Wikipedia.}

Wikipedia\footnote{url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability. At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages. Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.

\paragraph{Omniglot.}

The Omniglot website\footnote{url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages. We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us parallel data in a reasonably large number of languages. % The urls for these resources are:

% \begin{itemize}[noitemsep]
% \item \url{www.omniglot.com/language/phrases/*} 
% \item \url{www.omniglot.com/babel/*}
% \end{itemize}

\paragraph{ODIN.} \label{sec:odin}

ODIN (The Online Database of Interlinear Text) is a repository of
interlinear glossed texts (IGTs) extracted from scholarly documents
\cite{lewis2006odin,lewis2010odin}.  Compared to other resources, it
is notable for the breadth of languages included and the level of
linguistic annotation.  An IGT canonically consists of three lines:
(i) the source, a sentence in a target language, (ii) the gloss, an
analysis of each source element, and (iii) the translation,
done at the sentence level. The gloss line can additionally include a
number of linguistic terms, which means that the gloss 
%can more
%properly described as being written in a metalanguage, rather than a
is written in metalanguage rather than natural language.  In ODIN, translations are into English, and glosses
are written in an English-based metalanguage.  An accepted set of
guidelines are given by the Leipzig Glossing
Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}
where morphemes within words are separated by hyphens (or equal signs,
for clitics), and the same number of hyphens should appear in each
word of the source and gloss.

The data from ODIN poses the first obstacle to straightforwardly adopting Abney and Bird's data structure. 
The proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words. When we try to adapt IGTs into this format, we run into certain problems.  Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.  Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word). Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of \textit{word}, which undermines the central position of words in the proposed data structure.

Secondly, it is unclear how to represent the gloss.  Since the gloss line is not written in a natural language, we cannot treat it as a simple translation.  However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.  
% We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".  Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.  However, this loses information, since we no longer know which morpheme has which meaning.  To keep all information encoded in the IGT, we need to modify the \newcite{abney2011data}'s proposal.

Even though glosses in ODIN are based on English, having been
extracted from English-language documents, this is not true of IGTs in
general.  In particular, it is common for documentary linguists
working on indigenous languages of the Americas to provide glosses and
translations based on Spanish.  For this reason, we believe it would
be wise to specify the language used to produce the gloss.  Since it
is not quite the language itself, but a metalanguage, one solution
would be to use new language codes that make it clear both that a
metalanguage is being used, and also what natural language it is based
on.  The five-letter code \texttt{gloss} cannot be confused with any
code in any version of ISO 639 (with codes of length two to
four).  Following the convention that subvarieties of a language are
indicated with suffixes, we can append the code of the natural
language.  For example, glosses into English and
Spanish-based metalanguages would be given the codes
\texttt{gloss-eng} and \texttt{gloss-spa}, respectively.

@@

According to Ethnologue, there are 7105 living languages, and 147
living language families. Across all our data sources, we manage to
cover 1347 languages in 106 families, which represents 19.0\% of the
world's languages. To get a better idea of the kinds of languages
represented, we give a breakdown according to their EGIDS scores
(Expanded Graded Intergenerational Disruption Scale)
\cite{lewis2010assessing} in Figure \ref{fig:heatmap}. The values in
each cell have been colored according to proportion of languages
represented, with green indicating good coverage and red poor. It's
interesting to note that vigorous languages (6a) are poorly
represented across all data sources, and worse than more endangered
categories. In terms of language documentation, vigorous languages are
less urgent goals than those in categories 6b and up, but this
highlights an unexpected gap in linguistic resources.

@@

@@

### 2014-03-22 docs/computEL/intro.tex
edits to intro, related work, and data

At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{krauss1992crisis} estimated 50\% of the extant languages at the time were not being learned by new generations of speakers, risking extinction by the end of the century. Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.

Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, calling it the Human Language Project. Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing. The challenge was issued to the computational linguistics (CompLing) community, from the perspective that the language processing, machine learning, and data management tools\footnote{\textcolor{blue}{am I missing anything?}} well-known in CompLing must be brought to bear on the problems of documentary linguistics (DocLing), if we are to make any serious progress toward building such a resource. The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work of DocLing and increasing the efficiency of the language documentation process. \textcolor{blue}{[cite our AL work? is this too strong? also, run-on sentence. also, perhaps mention this workshop, which has precisely these goals]}

We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications for a universal corpus. We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format which is both in accordance with best practice archiving recommendations and readily accessible as training data for machine learning and for other computational methods. \textcolor{blue}{[kind of clunky]} Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}. To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.

To produce SeedLing, we have drawn on four web sources, described in section \ref{sec:sources}. To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation. We describe the steps required in detail in section \ref{sec:case_studies}, presenting each resource as a separate case study. By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus. To that end, many of the resources described in \ref{sec:related} are perhaps candidates for inclusion in the next stage of building a Universal Corpus.

We believe the resulting corpus, which at present covers 1347 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing. Following this claim, we use SeedLing in a test application (\ref{sec:cluster}): the task of language clustering. With no additional pre-processing required, we extract surface-level features (frequencies of character n-grams and words) for estimating the similarity of two languages. Unlike most previous approaches to the task, we make no use of resources curated for aims of linguistic typology (e.g. values of typological features as in WALS \cite{wals}, Swadesh word lists). Despite our approach being highly dependent on orthography, we achieve a reasonable clustering performance over the languages in our dataset, demonstrating SeedLing's utility in cross-linguistic research.

>>>>

At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{simons2011update} calculate that 37\% of the extant languages at the time were at various stages of losing transmisson to new generations of speakers.
Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.

Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, calling it the Human Language Project. Such a corpus would include data from all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing. The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulations and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource. The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work and increasing the efficiency of the language documentation process. 
%\textcolor{blue}{[cite our AL work? is this too strong? also, run-on sentence. also, perhaps mention this workshop, which has precisely these goals]}

We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications for a universal corpus. We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format which is both in accordance with best practice archiving recommendations and readily accessible as training data for machine learning and for other computational methods. Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}. To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.

To produce SeedLing, we have drawn on four web sources, described in section \ref{sec:sources}. To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation. We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate case study. By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus. To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a Universal Corpus.

We believe the resulting corpus, which at present covers 1347 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing. To test this claim, we use SeedLing in a sample application (section \ref{sec:cluster}): the task of language clustering. With no additional pre-processing required, we extract surface-level features (frequencies of character n-grams and words) for estimating the similarity of two languages. Unlike most previous approaches to the task, we make no use of resources curated for aims of linguistic typology (e.g. values of typological features as in WALS \cite{wals}, Swadesh word lists). Despite our approach being highly dependent on orthography, we achieve a reasonable clustering performance over the languages in our dataset, demonstrating SeedLing's utility in cross-linguistic research.

### 2014-03-22 docs/computEL/paper.tex
edits to intro, related work, and data

\begin{document}
\maketitle
\begin{abstract}
  A bread-coverage corpus such as the Human Language Project
  envisioned by \newcite{abney2010universal} would be a powerful
  resource for the study of endangered languages.
%, \bluenote{keep? from
 %   the perspective of both the documentary and the computational
  %  linguist}. 
Existing corpora are limited in the range of languages
  covered, in standardisation, or in machine-readability. In this
  paper we present SeedLing, a seed corpus for the Human Language
  Project. We first survey existing efforts to compile
  cross-linguistic resources, then describe our own approach. To build
  the foundation text for a universal corpus, we crawl and clean
  texts from several web sources that contain data from a large number
  of languages, and convert them into a standardised form consistent
  with the guidelines set out by \shortcite{abney2010universal}. The
  resulting corpus is more easily-accessible and machine-readable than
  any of the underlying data sources, and represents a significant
  base corpus for researchers to draw on and add to in the future. To show the utility of SeedLing for cross-lingual computational research, we use our data in the test application of clustering similar languages.
\end{abstract}

>>>>

\begin{document}
\maketitle
\begin{abstract}
  A bread-coverage corpus such as the Human Language Project
  envisioned by \newcite{abney2010universal} would be a powerful
  resource for the study of endangered languages.
%, \bluenote{keep? from
 %   the perspective of both the documentary and the computational
  %  linguist}. 
Existing corpora are limited in the range of languages
  covered, in standardisation, or in machine-readability. In this
  paper we present SeedLing, a seed corpus for the Human Language
  Project. We first survey existing efforts to compile
  cross-linguistic resources, then describe our own approach. To build
  the foundation text for a universal corpus, we crawl and clean
  texts from several web sources that contain data from a large number
  of languages, and convert them into a standardised form consistent
  with the guidelines of \newcite{abney2011data}. The
  resulting corpus is more easily-accessible and machine-readable than
  any of the underlying data sources, and represents a significant
  base corpus for researchers to draw on and add to in the future. To show the utility of SeedLing for cross-lingual computational research, we use our data in the test application of clustering languages.
\end{abstract}

### 2014-03-22 docs/computEL/relwork.tex
edits to intro, related work, and data

\subsection{Traditional Archives}

Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.  Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access to their data.

Some archives aim for a universal scope, including languages from all parts of the world.  Of particular note are:

The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen), among many other projects.
Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale).
The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, London, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).

Regional archives include:

AILLA (Archive of Indigenous Languages of Latin America), at the University of Texas at Austin. (Not in OLAC)
AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies) (Not in OLAC)
California Language Archive, managed by the Survey of California and Other Indian Languages
ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks.
PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures)

However, there are two main problems common to all of the above data sources.  Firstly, the data is not always machine readable.  For instance, in their own words, ANLA consists of "mostly paper records".  Even where the data is available digitally, these often take the form of scanned images, or audio files.  While both can provide invaluable information, they are extremely difficult to process with a computer, requiring an impractical level of pre-processing of the image or audio, before linguistic analysis can begin.  Even textual data, which avoids these issues, may not be available in a machine-readable form, being stored as pdfs or

Secondly, when data is machine readable, the format can vary wildly.  This makes automated processing difficult, especially if one is not aware of the details of each project.  Even when metadata standards and encodings agree, there can be idiosyncractic markup or non-linguistic information, such as labels for speakers in the transcript of a conversation.

We can see that there is still much work to be done by individual researchers in digitising and standardising linguistic data, and it is outside of the scope of this paper to attempt this for the above archives.  Guidelines for producing new materials are available from the E-MELD project (Electronic Metastructure for Endangered Languages Data), which specifically aimed to deal with the expanding number of standards for linguistic data.  It gives best practice recommendations, illustrated with eleven case studies, and provides input tools which link to the GOLD ontology language, and the OLAC metadata set.  Further recommendations are given by \newcite{bird2003port}, who describes seven dimensions along which the portability of linguistic data can vary. Various tools are available from The Language Archive at the Max Planck Institute for Psycholinguistics.

Many of these archives are part of the Open Language Archive Community (OLAC), a subcommunity of the Open Archives Initiative.  OLAC maintains a metadata standard, based on the 15-element Dublin Core, which allows a user to search through all participating archives in a unified fashion.  However, centralising access to disparate resources, while of course extremely helpful, does not solve the problem of inconsistent standards.  Indeed, it can become difficult even to answer simple questions like "how many languages are represented?"

We do of course acknowledge that such resources have their purpose - for instance, there can be plenty of material to work with for a researcher who is looking to find data on a specific language, who is willing to put in a little time browsing through different catalogues, and who only needs human-readable data.  However, for large-scale machine processing, they leave much to be desired.

\subsection{Generic corpora collections}

There are some corpora collections which do not take endangered languages to be a main aim, but which nonetheless cover an increasing number of languages.

MetaShare (Multilingual Europe Technology Alliance) provides data in a little over 100 languages. While language codes are used, they have not been standardised, so that multiple codes are used for the same language.

Linguistic Data Consortium (LDC) and the European Language Resources Association (ELRA) both offer data in multiple languages.  However, while large in size, they cover only a limited number of languages.  Furthermore, the corpora they contain are stored separately, making it difficult to access data according to language.

It seems likely that resources which do not specifically aim to cover endangered languages will continue to be skewed heavily towards a handful of high-resource languages.

\subsection{Parallel Corpora}

The Machine Translation community has assembled a number of parallel corpora, which are crucial for statistical machine translation. The OPUS corpus \cite{tiedemann2012opus} subsumes a number of other well-known parallel corpora, such as Europarl, and covers documents from 350 languages, with various language pairs.  However, not all of the data has been manually checked, which introduces noise.

\subsection{Web Corpora}

@@

HC Corpora currently provides download of corpora in 68 different language varieties, which vary in size from 2M to 150M words. The corpora are thus of a respectable size, but only 1% of the world's languages are represented.  A further difficulty is that languages are named, without the corresponding ISO codes. This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.

The Leipzig Corpora Collection (LCC)\footnote{http://corpora.uni-leipzig.de} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230. The corpora are large, readily available, in plain text, and labelled with ISO codes.

@@

While the above efforts are commendable, and allow cross-linguistic analysis to a certain extent, there is still a long way to go to producing a universal corpus.

\subsection{Cross-linguistic projects}

Responding to the call to document and preserve the world's languages, highly cross-linguistic projects have sprung up, striving towards the aim of universality.  Of particular note are the Endangered Languages Project, and the Rosetta Project.

However, as much as these projects are to be praised for their commitment to universality, it is difficult to use their data to perform large-scale NLP.  A researcher looking for a data on a specific language can easily navigate to the language in question, and download the available resources - but trying to access all the data can be awkward.  While the sleek online interface used by the Endangered Languages Project may make it more accessible to those unfamiliar with language processing, it is not ideal for a computational linguistics researcher.

>>>>

\paragraph{Traditional archives.}
Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.  Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access to their data.

Some archives aim for a universal scope, including languages from all parts of the world.  Of particular note are: The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen) and many other projects;
Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale);
The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).

Regional archives include: AILLA (Archive of Indigenous Languages of Latin America), University of Texas at Austin;
AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies);
California Language Archive, managed by the Survey of California and Other Indian Languages;
ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks;
PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures).

However, there are two main problems common to all of the above data sources.  Firstly, the data is not always machine readable.  For instance, in their own words, ANLA consists of ``mostly paper records''.  Even where the data is available digitally, these often take the form of scanned images or audio files.  While both can provide invaluable information, they are extremely difficult to process with a computer, requiring an impractical level of image or video pre-processing before linguistic analysis can begin.  Even textual data, which avoids these issues, may not be available in a machine-readable form, being stored as pdfs or other opaque formats.
Secondly, when data is machine readable, the format can vary wildly.  This makes automated processing difficult, especially if one is not aware of the details of each project.  Even when metadata standards and encodings agree, there can be idiosyncractic markup or non-linguistic information, such as labels for speakers in the transcript of a conversation.

We can see that there is still much work to be done by individual researchers in digitising and standardising linguistic data, and it is outside of the scope of this paper to attempt this for the above archives.  Guidelines for producing new materials are available from the E-MELD project (Electronic Metastructure for Endangered Languages Data), which specifically aimed to deal with the expanding number of standards for linguistic data.  It gives best practice recommendations, illustrated with eleven case studies, and provides input tools which link to the GOLD ontology language, and the OLAC metadata set.  Further recommendations are given by \newcite{bird2003port}, who describe seven dimensions along which the portability of linguistic data can vary. Various tools are available from The Language Archive at the Max Planck Institute for Psycholinguistics.

Many of these archives are part of the Open Language Archive Community (OLAC), a subcommunity of the Open Archives Initiative.  OLAC maintains a metadata standard, based on the 15-element Dublin Core, which allows a user to search through all participating archives in a unified fashion.  However, centralising access to disparate resources, while of course extremely helpful, does not solve the problem of inconsistent standards.  Indeed, it can be difficult even to answer simple questions like ``how many languages are represented?''

Such resources are invaluable for many purposes
for instance, there can be plenty of material to work with for a
researcher who is looking to find data on a specific language, who is
willing to put in a little time browsing through different catalogues,
and who only needs human-readable data.  However, for large-scale
machine processing, they leave much to be desired.

\paragraph{Generic corpus collections.}
Some corpus collections exist which do not focus on endangered languages, but which nonetheless cover an increasing number of languages.

MetaShare (Multilingual Europe Technology Alliance) provides data in a little over 100 languages. 
While language codes are used, they have not been standardised, so that multiple codes are used for the same language.  Linguistic Data Consortium (LDC) and the European Language Resources Association (ELRA) both offer data in multiple languages.  However, while large in size, they cover only a limited number of languages.  Furthermore, the corpora they contain are stored separately, making it difficult to access data according to language.

% It seems likely that resources which do not specifically aim to cover endangered languages will continue to be skewed heavily towards a handful of high-resource languages.

\paragraph{Parallel corpora.}

The Machine Translation community has assembled a number of parallel corpora, which are crucial for statistical machine translation. The OPUS corpus \cite{tiedemann2012opus} subsumes a number of other well-known parallel corpora, such as Europarl, and covers documents from 350 languages, with various language pairs.

%However, not all of the data has been manually checked, which introduces noise.

\paragraph{Web corpora.}

@@

HC Corpora currently provides download of corpora in 68 different language varieties, which vary in size from 2M to 150M words. The corpora are thus of a respectable size, but only 1\% of the world's languages are represented.  A further difficulty is that languages are named, without the corresponding ISO codes. This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.

The Leipzig Corpora Collection (LCC)\footnote{http://corpora.uni-leipzig.de} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230. The corpora are large, readily available, in plain text, and labelled with ISO language codes.

@@

% While the above efforts are commendable, and allow cross-linguistic analysis to a certain extent, there is still a long way to go to producing a universal corpus.

\paragraph{Cross-linguistic projects.}

Responding to the call to document and preserve the world's languages, highly cross-linguistic projects have sprung up, striving towards the aim of universality.  Of particular note are the Endangered Languages Project, and the Rosetta Project. These projects are to be praised for their commitment to universality, but in their current forms it is difficult to use their data to perform large-scale NLP.

% A researcher looking for a data on a specific language can easily navigate to the language in question, and download the available resources - but trying to access all the data can be awkward.  While the sleek online interface used by the Endangered Languages Project may make it more accessible to those unfamiliar with language processing, it is not ideal for a computational linguistics researcher.
